---
title: "Prepare phenotype data"
author: "Luis Fdo Delgado"
date: "2022-03-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(tidyverse);
library(genomicMateSelectR);
library(magrittr)
```


## Read DB data


```{r}
indata <- read.csv(here::here("data","phenotype.csv"),
                 na.strings = c("#VALUE!",NA,".",""," ","-","\""),
                 stringsAsFactors = F, skip=3)

meta <- read.csv(here::here("data","metadata.csv"),
               na.strings = c("#VALUE!",NA,".",""," ","-","\""),
               stringsAsFactors = F, skip=2)

meta %<>%
     dplyr::rename(programName = breedingProgramName,
                   programDescription = breedingProgramDescription,
                   programDbId = breedingProgramDbId)
indata <- dplyr::left_join(indata,meta)

indata %<>% # <- + %>% 
     dplyr::filter(observationLevel=="plot")

dbdata <- indata

```

## Check experimental designs

Checklist: Are the data plot-basis, plant-basis or a mixture? If plant-basis data are present, should they be converted to plot-basis for further analysis?

```{r}
dbdata %>% count(observationLevel) # number of observation
```
*Checklist:* What experimental designs are present? How are they represented by the variables in the dataset? Are all designs consistent with your expectations, for example relative to the reported “trialType,” “studyName” and/or “studyDesign?”


```{r}
library(gt)
dbdata %>% 
     count(studyName,trialType, studyDesign, numberBlocks,numberReps,entryType) %>% 
     spread(entryType,n) 

```

Summary table above shows:

1. **trialType** and **studyDesign** cannot be 100% relied upon, at least not here.
2. The only trial actually listed as having `studyDesign=="Augmented"` does not have “check” vs. “test” distinguished in the “entryType.”
3. A `trialType=="Clonal Evaluation"` with `studyDesign=="RCBD"` but actually only 1 replication.

Next, I’ll check if the **replicate** and **blockNumber** columns reliably distinguish complete and incomplete blocks in the data.


```{r}
dbdata %>% 
     group_by(studyName) %>% 
     summarize(N_replicate=length(unique(replicate)),
               N_blockNumber=length(unique(blockNumber)))
```

Here, I notice that except 1 trial (**19.GS.C1.C2.C3.AYT.42.UB**) has the same number of reps and blocks.

The question is, are complete replications of the experiment indicated by **replicate** and incomplete sub-blocks represented by **blockNumber**

```{r}
dbdata %>% 
     group_by(studyName) %>% 
     summarize(N_replicate=length(unique(replicate)),
               N_blockNumber=length(unique(blockNumber)),
               doRepsEqualBlocks=all(replicate==blockNumber))

```

So for 1 trial, there are 3 complete blocks, no sub-blocks. For 6 trials, there are 2 complete replications and nested sub-blocks represented by the **blockNumber** variable. For 2 trials, there are only incomplete blocks.

Next, I decided to check that the **replicate** column definitely means complete blocks. The below might look a bit complicated, but I basically merge two summaries: (1) he overall number of accessions per trial, and (2) the average number of accessions per replicate per trial.

```{r}
# the overall number of accessions per trial
dbdata %>% 
     group_by(studyName) %>% 
     summarize(N_accession=length(unique(germplasmName)))
```

```{r}
dbdata %>% 
     group_by(studyName,replicate) %>% 
     summarize(N_accession=length(unique(germplasmName))) %>% 
     group_by(studyName) %>% 
     summarize(avgAccessionsPerReplicate=ceiling(mean(N_accession)))
```

```{r}
# the overall number of accessions per trial
dbdata %>% 
     group_by(studyName) %>% 
     summarize(N_accession=length(unique(germplasmName))) %>% 
     # the average number of accessions per replicate per trial
     left_join(dbdata %>% 
                    group_by(studyName,replicate) %>% 
                    summarize(N_accession=length(unique(germplasmName))) %>% 
                    group_by(studyName) %>% 
                    summarize(avgAccessionsPerReplicate=ceiling(mean(N_accession))))
```



The numbers are very similar for all trials, indicating complete blocks.

One more: look at the min, mean and max number of accessions per **blockNumber.**

```{r}
dbdata %>% 
     group_by(studyName,replicate,blockNumber) %>% 
     summarize(N_accession=length(unique(germplasmName))) %>% ungroup() %>% 
     group_by(studyName) %>% 
     summarize(minAccessionsPerBlock=ceiling(min(N_accession)),
               avgAccessionsPerBlock=ceiling(mean(N_accession)),
               maxAccessionsPerBlock=ceiling(max(N_accession)))
```

```{r}
dbdata %>% 
     group_by(studyName) %>% 
     summarize(N_accession=length(unique(germplasmName))) %>% 
     left_join(dbdata %>% 
                    group_by(studyName,replicate,blockNumber) %>% 
                    summarize(N_accession=length(unique(germplasmName))) %>% ungroup() %>% 
                    group_by(studyName) %>% 
                    summarize(minAccessionsPerBlock=ceiling(min(N_accession)),
                              avgAccessionsPerBlock=ceiling(mean(N_accession)),
                              maxAccessionsPerBlock=ceiling(max(N_accession))))
```
AYT trial does not have incomplete blocks… only complete blocks.

Now I will ad hoc create two variables (CompleteBlocks and IncompleteBlocks), indicating (TRUE/FALSE) whether to model using the replicate and/or blockNumber variable.

I also like to create explicitly nested design variables (yearInLoc, trialInLocYr, repInTrial, blockInRep).

```{r}
dbdata %<>% 
     group_by(studyName) %>% 
     summarize(N_replicate=length(unique(replicate)),
               N_blockNumber=length(unique(blockNumber)),
               doRepsEqualBlocks=all(replicate==blockNumber)) %>% 
     ungroup() %>% 
     mutate(CompleteBlocks=ifelse(N_replicate>1,TRUE,FALSE),
            IncompleteBlocks=ifelse(N_blockNumber>1 & !doRepsEqualBlocks,TRUE,FALSE)) %>% 
     left_join(dbdata) %>% 
     mutate(yearInLoc=paste0(programName,"_",locationName,"_",studyYear),
            trialInLocYr=paste0(yearInLoc,"_",studyName),
            repInTrial=paste0(trialInLocYr,"_",replicate),
            blockInRep=paste0(repInTrial,"_",blockNumber))
```
Just to check:

```{r}
dbdata %>% 
     count(studyName,CompleteBlocks,IncompleteBlocks) %>% 
     left_join(dbdata %>% 
                    group_by(studyName) %>% 
                    summarize(nRepInTrial=length(unique(repInTrial)),
                              nBlockInRep=length(unique(blockInRep))))
```

# Traits and Trait Abbreviations

Cassavabase downloads use very long column-names corresponding to the [full trait-ontology name](https://cropontology.org/ontology/CO_334). For convenience, I replace these names with abbreviations, documented here. For eventual upload of analysis results, names will need to be restored to ontology terms.

I also use this opportunity to subselect traits.

```{r}
traitabbrevs<-tribble(~TraitAbbrev,~TraitName,
        "CMD1S","cassava.mosaic.disease.severity.1.month.evaluation.CO_334.0000191",
        "CMD3S","cassava.mosaic.disease.severity.3.month.evaluation.CO_334.0000192",
        "CMD6S","cassava.mosaic.disease.severity.6.month.evaluation.CO_334.0000194",
        "DM","dry.matter.content.percentage.CO_334.0000092",
        "RTWT","fresh.storage.root.weight.per.plot.CO_334.0000012",
        "NOHAV","plant.stands.harvested.counting.CO_334.0000010")
#rmarkdown::paged_table()
```



```{r}
dbdata <- renameAndSelectCols(traitabbrevs,
                            indata=dbdata,
                            customColsToKeep = c("observationUnitName",
                                                 "CompleteBlocks",
                                                 "IncompleteBlocks",
                                                 "yearInLoc",
                                                 "trialInLocYr",
                                                 "repInTrial","blockInRep"))
#> Joining, by = "TraitName"

```

# Dry matter boxplot

```{r fig.align='center', fig.height=7, fig.width= 10}
dbdata %>% ggplot(aes(reorder(x = studyName, DM), y = DM, fill = studyName))+
  geom_boxplot(show.legend = F) +
  geom_jitter(width = 0.2, show.legend = F, alpha = 0.2) +
  labs(x = NULL) +
  theme_light() +
  theme(axis.text.x = element_text(hjust = 1 , angle = 90),
        text = element_text(size = 15))
  
```

# QC Trait Values

At this point in the pipeline, we should check the all trait values are in allowable ranges. Different ways to approach this. Feel free to make some plots of your data!

The database also has mechanisms to ensure trait values are only within allowable ranges.

Nevertheless, as a habit, I have an simple ad hoc approach to this:

```{r}
# comment out the traits not present in this dataset
dbdata<-dbdata %>% 
     dplyr::mutate(CMD1S=ifelse(CMD1S<1 | CMD1S>5,NA,CMD1S),
                   CMD3S=ifelse(CMD3S<1 | CMD3S>5,NA,CMD3S),
                   # CMD6S=ifelse(CMD6S<1 | CMD6S>5,NA,CMD6S), 
                   # CMD9S=ifelse(CMD9S<1 | CMD9S>5,NA,CMD9S),
                   # CGM=ifelse(CGM<1 | CGM>5,NA,CGM),
                   # CGMS1=ifelse(CGMS1<1 | CGMS1>5,NA,CGMS1),
                   # CGMS2=ifelse(CGMS2<1 | CGMS2>5,NA,CGMS2),
                   DM=ifelse(DM>100 | DM<=0,NA,DM),
                   RTWT=ifelse(RTWT==0 | NOHAV==0 | is.na(NOHAV),NA,RTWT),
                   # SHTWT=ifelse(SHTWT==0 | NOHAV==0 | is.na(NOHAV),NA,SHTWT),
                   # RTNO=ifelse(RTNO==0 | NOHAV==0 | is.na(NOHAV),NA,RTNO),
                   NOHAV=ifelse(NOHAV==0,NA,NOHAV),
                   NOHAV=ifelse(NOHAV>42,NA,NOHAV)
                   # RTNO=ifelse(!RTNO %in% 1:10000,NA,RTNO)
     )
```

# Post-QC: composite traits

Now that component traits are QC’d, it’s time to compute any composite traits.

By composite traits, I mean traits computed from combinations of other traits.

Examples for cassava: season-wide mean disease severity, harvest index, and fresh root yield.

```{r}
# [NEW AS OF APRIL 2021]
## VERSION with vs. without CBSD
## Impervious to particular timepoints between 1, 3, 6 and 9 scores

# Without CBSD (West Africa)
dbdata<-dbdata %>% 
  mutate(MCMDS=rowMeans(.[,colnames(.) %in% c("CMD1S","CMD3S","CMD6S","CMD9S")], na.rm = T)) %>% 
  select(-any_of(c("CMD1S","CMD3S","CMD6S","CMD9S")))

# With CBSD (East Africa)
# dbdata<-dbdata %>% 
#   mutate(MCMDS=rowMeans(.[,colnames(.) %in% c("CMD1S","CMD3S","CMD6S","CMD9S")], na.rm = T),
#          MCBSDS=rowMeans(.[,colnames(.) %in% c("CBSD1S","CBSD3S","CBSD6S","CBSD9S")], na.rm = T)) %>% 
#   select(-any_of(c("CMD1S","CMD3S","CMD6S","CMD9S","CBSD1S","CBSD3S","CBSD6S","CBSD9S")))
```

# Fresh root yield (FYLD)

**RTWT** (fresh root weight per plot in kg) –> **FYLD** (fresh root yield in tons per hectare)

```{r}
dbdata %>% 
     count(studyYear,studyName,studyDesign,plotWidth,plotLength,plantsPerPlot) %>% 
     mutate(plotArea=plotWidth*plotLength) %>% 
     gt() %>% tab_options(table.font.size = pct(67))
```
In the example trial data, the plantsPerPlot meta-data field is empty. To my knowledge, no meta-data field is available in BreedBase to represent a net-plot harvest.

**RECOMMEND INPUTING plantsPerPlot meta-data to cassavabase for your breeding program!**

Firstly noting that the trial 19geneticgainUB actually does not have phenotypes (for any trait). It will be excluded downstream. (~~I might find a substitute genetic gain trial, from an earlier year, for the sake of this example~~)

To decide what the real MaxHarvestedPlantsPerPlot and plantsPerPlot were likely to have been, I make two plots below and also compute the maximum NOHAV for each trial.

```{r fig.align='center', fig.height=7, fig.width= 10}
dbdata %>% 
     ggplot(.,aes(x=NOHAV, fill=studyName)) + geom_density(alpha=0.75)
```

Maybe clearer to make a boxplot?

```{r fig.align='center', fig.height=7, fig.width= 10}
dbdata %>% 
     # plot area in meters squared
     mutate(plotArea=plotWidth*plotLength) %>% 
     ggplot(.,aes(x=plotArea,y=NOHAV, fill=studyName)) + 
     geom_boxplot() + theme(axis.text.x = element_blank())
#> Warning: Removed 921 rows containing non-finite values
#> (stat_boxplot).

dbdata %>% 
     # plot area in meters squared
     mutate(plotArea=plotWidth*plotLength) %>% select(plotArea) %>% max()
```


```{r}
plantsPerPlot_choices <- dbdata %>% 
     distinct(studyYear,studyName,plotWidth,plotLength,plantsPerPlot) %>% 
     left_join(dbdata %>% 
                    group_by(studyName) %>% 
                    summarize(MaxNOHAV=max(NOHAV, na.rm=T))) %>% 
          # plot area in meters squared
     mutate(plotArea=plotWidth*plotLength,
            # Number of plants per plot
            plantsPerPlot=MaxNOHAV,
            plantsPerPlot=ifelse(studyName=="19.GS.C2.UYT.36.setA.UB",20,plantsPerPlot)) %>% 
     # exclude the empty genetic gain trial
     filter(studyName!="19geneticgainUB") %>% 
     select(studyName,plotArea,MaxNOHAV,plantsPerPlot)

plantsPerPlot_choices 
```
**As a data generator, in-house at each breeding program, no reason not to get the correct answer and repair the metadata on the database!**

```{r fig.align='center', fig.height=7, fig.width= 10}
dbdata %<>%
     select(-plantsPerPlot) %>% 
     # join plantsPerPlot_choices to the trial data
     left_join(plantsPerPlot_choices) %>% 
     # compute fresh root yield (FYLD) in tons per hectare
     mutate(PlantSpacing=plotArea/plantsPerPlot,
            FYLD=RTWT/(plantsPerPlot*PlantSpacing)*10) # FYLD (fresh root yield in tons per hectare)


dbdata %>% ggplot(.,aes(x=FYLD,fill=studyName)) + geom_density(alpha=0.75) 
```

**Log-transform yield traits: this is a habit based on experience. Linear mixed-models should have normally distributed homoskedastic residuals, if they don’t log-transform the response variable often helps. For FYLD and related traits, I always log-transform.**

```{r fig.align='center', fig.height=7, fig.width= 10}
dbdata %<>% mutate(DYLD=FYLD*(DM/100), # DYLD (dry root yield in tons per hectare)
            logFYLD=log(FYLD),
            logDYLD=log(DYLD),
            PropNOHAV=NOHAV/plantsPerPlot) 
# remove non transformed / per-plot (instead of per area) traits
dbdata %<>% select(-RTWT,-FYLD,-DYLD)
dbdata %>% ggplot(.,aes(x=logFYLD,fill=studyName)) + geom_density(alpha=0.75) 
```

Debatable whether this is better. Let’s not dwell on it. Onward!

**SUGGESTION**: For individuals working this manual, consider making different, or no transformations as you see fit, with your own data. Even better, set-up a direct comparison of results with- vs. without-transformation.*

# Save “cleaned” phenotypes

```{r}
saveRDS(dbdata,file=here::here("output","phenotypes_cleaned.rds"))
```











